Continuous gravitational waves in the lab: recovering audio signals with a table-top optical microphone

### 2020-10-01 paper/ifo-conclusions.tex
Deeksha's comments - last few sections

@@

In this paper we present an educational demonstration in which a table-top Michelson interferometer serves as an analogue to a gravitational-wave detector. 
Vibrations produced by gravitational-wave signals are simulated with sound recordings played though a speaker fixed to the back of one of the interferometer mirrors. 
We consider quasi-monochromatic signals as an analogue to continuous wave signals as well as short segments of speech and music which are (more distant) analogues of unmodelled gravitational-wave bursts, e.g. from supernovae. 
% This serves to demonstrate signal processing techniques.

In Sections~\ref{sec:ifo} and ~\ref{sec:single_tone} we demonstrate the recovery of signals with constant and slowly wandering frequency. 
A webcam sampling at $30\,{\rm Hz}$ is used to track the interference pattern and the injected signal is recovered successfully with the Viterbi algorithm. 
These sections directly demonstrate the continuous wave search techniques used in gravitational-wave data analysis. 
%Using a webcam to track the interference pattern, we inject a wandering frequency signal into the system with a speaker attached to the back of one of mirrors of the interferometer. We then demonstrate how to recover this signal using the Viterbi algorithm, just as is done in continuous gravitational wave searches.

In Section~\ref{sec:optical_microphone} we demonstrate recovery of more complex signals of music and speech. 
Substituting the webcam for the increased frequency response of a photodiode read by a Raspberry Pi, we transform the interferometer into an optical microphone.
We demonstrate that the optical microphone is capable of playing back simple recordings well but produces unintelligible recordings of speech. 
Mains hum at $50\,{\rm Hz}$ and its harmonics dominates the photodiode output. 
This demonstration offers students the opportunity to explore a variety of filtering and speech enhancement techniques. 
Further avenues of exploration include a more complete model of the optical microphone transfer function, which may lead to improvements in intelligibility.

The demonstration presented here has good potential to be adapted for use in the undergraduate laboratory. 
It can be used to demonstrate topics of interest to physicists and electrical engineers including: interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement. 
Given the increased excitement and public interest in gravitational-wave science since the detection in 2015, this table-top demonstration presented here can also be adapted as a tool to explain gravitational-wave research developments for a wider non-specialist audience. 
The future will bring many more detections, including further observations of transient signals from binary black hole and binary neutron star mergers, as well as the anticipated first detections of other classes of signal, including continuous waves.

>>>>

In this paper, we use a table-top Michelson interferometer as an analogue to a gravitational-wave detector and demonstrate a range of signal processing techniques used within the GW community and more broadly. Gravitational-wave signals are simulated by playing sound recordings through a speaker which is fixed to the back of one of the arm mirrors. We consider quasi-monochromatic signals as an analogue to continuous wave signals as well as short segments of speech and music for more general interest. These recordings of complex audio could also serve as more distant analogues to un-modelled gravitational-wave bursts, e.g. from supernovae.

We demonstrate the recovery of a signal with constant frequency and signal with slowly wandering frequency in Sections~\ref{sec:ifo} and ~\ref{sec:single_tone}, respectively. We use a basic webcam with a sampling rate of $30\,{\rm Hz}$ to track the interference pattern. For the wandering frequency we successfully recover the injected signal using the Viterbi algorithm. These sections directly demonstrate the techniques used in continuous gravitational-wave searches.

In Section~\ref{sec:optical_microphone}, we demonstrate the recovery of more complex signals of music and speech. By substituting the webcam with a photodiode read by a Raspberry Pi, we transform the interferometer into an optical microphone by greatly increasing the sampling rate. We demonstrate that the optical microphone is capable of playing back simple recordings well but produces unintelligible recordings of speech. The mains signal at $50\,{\rm Hz}$ and its harmonics dominate the photodiode output. To recover the signal from under this noise we apply a hierarchy of filters and statistical techniques from signals processing, including a Butterworth band-pass filter and a notch filter, to a Wiener filter and the logMMSE estimator. We find that the logMMSE estimator works the best from those tried but still doesn’t enhance the speech to an intelligible level. This demonstration allows students to explore a variety of filtering and speech enhancement techniques. Further avenues of exploration include even more advanced processing techniques and a full characterisation of the transfer function of the system, with the goal to lead to improvements in intelligibility.

The demonstration presented here has great potential to be adapted for use in the undergraduate laboratory. It can be used to demonstrate topics such as interferometer physics; gravitational-wave detection, searches, and analysis; and signal processing with filters and speech enhancement techniques to physics and electrical engineering students. Given the increased level of excitement and public interest in the field of gravitational-wave astronomy in recent years, this table-top demonstration can be used as a tool for explaining gravitational-wave research to a wider, non-specialist audience. The future will bring many more detections, including further observations of transient signals from binary black hole and binary neutron star mergers, as well as the anticipated first detections of other classes of signal, including continuous waves, to which this demonstration provides some charming insights.

@@

%Using a webcam to track the interference pattern, we inject a wandering frequency signal into the system with a speaker attached to the back of one of mirrors of the interferometer. We then demonstrate how to recover this signal using the Viterbi algorithm, just as is done in continuous gravitational wave searches.

### 2020-10-01 paper/ifo-futureWork.tex
Deeksha's comments - last few sections

\documentclass[paper-main.tex]{subfiles}

\begin{document}

\begin{figure*}
	\includegraphics[height=.99\textwidth, angle=-90]{figures/pipeline.pdf}
	\caption{
Flowchart for the optical microphone transfer function. 
The stage inside the red-dash box is analysed in Appendix~\ref{app:intensity_derivation}. Understanding the other stages is one aim of future work. 
}
	\label{fig:pipeline_highlighted}
\end{figure*}

The optical microphone experiment describe here can be used to convey a variety of topics as a teaching tool. 
Here we consider a selection of further extensions to the experiment touching on noise isolation, transfer function analysis, speech recognition, and control systems.

An initial improvement is to better isolate the experiment from interference, such as the mains supply. 
The Raspberry Pi is USB powered and requires low power so it could be run off a commercial external phone battery. 
Everything else in the circuit, including the photodiode, is connected to the Pi. 
The laser is DC powered through a converter, so should not introduce mains harmonics itself. 
Transferring the circuit from a breadboard to a printed circuit board will also reduce interference~\cite{elfekey2013design}

%With the interferometer in a dark room and under a box to better isolate from any ambient light or air currents, this would hopefully remove (or better separate) all sources of the 50Hz mains hum from the optical microphone.

A second improvement is to measure the full model of the transfer function of the system, providing more information about how the audio signal couples with the interferometer.
% commented out whilst James is checking this
%With perfect knowledge of the transfer function, signal recovery is limited by the signal-to-noise ratio as an optimal matched filter could be constructed.
%\jam{[check with Rob et al. about the veracity of this]}
The transfer function starts from the voltage sent to the speaker and ends with the voltage recorded by the Pi. It includes any non-linearities in the speaker itself, the speaker-mirror coupling, the path length to intensity relation (Appendix~\ref{app:intensity_derivation}), and the photodiode reading. 
A schematic of the full optical microphone transfer function is shown in Figure~\ref{fig:pipeline_highlighted}.

A voluminous literature exists about using a hidden Markov model (HMM) trained against phonemes to recognize speech~\cite{HMM_english}. 
Interestingly, this connects back to the Viterbi algorithm in Section IV, which is also underpinned by a hidden Markov model, albeit of a different type. 
Alternatively, machine learning solutions exist throughout the field that are able to compete~\cite{SEGAN} with statistical techniques like the logMMSE estimator.

Other traditional signal processing methods for extracting spectrogram like wavelet transform (WT) \citep{nason1995stationary} can also be implemented and compared with HMM. WT provides both time and frequency information, making it easier to pinpoint the origin of noise respect to time. 
In Refs.~\cite{tufekci2000feature,agbinya1996discrete}, WT was also proposed for speech recognition.

Finally, we can also consider how to improve the stability of the optical microphone components.
A feedback loop control system can be used to suppress any unwanted movement of the interferometer components~\citep{abbott2017exploring, Sekiguchi:2016bmv, verhoeven2009robust}.

>>>>

\documentclass[paper-main.tex]{subfiles}
\begin{document}

The optical microphone experiment described here can be used to demonstrate and teach a variety of topics including basic optics and photodiode circuits, to signals processing and speech enhancement. Here we consider further extensions and improvements to the experiment such as noise isolation, transfer function analysis, speech recognition, and control systems.

An initial improvement could be to better isolate the experiment, especially from the sources of mains noise. As the Raspberry Pi demands little power and connects to USB it could instead be powered with a commercial battery rather than the mains supply. All other components of the circuit, including the photodiode are connected to the Pi. The laser is DC powered through a converter, so should not couple noise from the mains into the interferometer. Transferring the circuit from a breadboard to a printed circuit board can also reduce interference~\cite{elfekey2013design}.
%With the interferometer in a dark room and under a box to better isolate from any ambient light or air currents, this would hopefully remove (or better separate) all sources of the 50Hz mains hum from the optical microphone.

We can consider improving the stability of the optical microphone components. A feedback loop control system can be used to suppress any unwanted movement of the interferometer components~\citep{abbott2017exploring, Sekiguchi:2016bmv, verhoeven2009robust}.

We can also benefit from fully characterizing the transfer function of the system as it provides the exact information of how audio signal couples through the interferometer. The total transfer function would start with the voltage being sent to the speaker and end with the voltage recorded by the Pi. It also would include any inherent non-linearities in the speaker, the speaker-mirror coupling, the path length to intensity relation (Appendix~\ref{app:intensity_derivation}), and the photodiode. The schematic in Fig.~\ref{fig:pipeline_highlighted} shows the signal flow through the system.
% commented out whilst James is checking this
% jam: I'm pretty confident this is what Rob said. 
%With perfect knowledge of the transfer function, signal recovery is limited by the signal-to-noise ratio as an optimal matched filter could be constructed.
%\jam{[check with Rob et al. about the veracity of this]}

A voluminous literature exists on using a hidden Markov model (HMM) trained on phonemes to recognize speech~\cite{HMM_english}. These could perform the final stage of speech enhancement for the optical microphone. Interestingly, this connects back to the Viterbi algorithm in Section~\ref{sec:viterbi_wandering}, which is also underpinned by a HMM, albeit of a different type. Alternatively, machine learning solutions exist throughout the field that are able to compete with statistical techniques such as the logMMSE estimator~\cite{SEGAN}.

We could also use more traditional techniques such as wavelet transform (WT) \citep{nason1995stationary} to extract the signal from noise, and compare these with the above methods. WT provides both time and frequency information, making it easier to pinpoint the origin of noise with respect to time. The authors in Refs.~\cite{tufekci2000feature,agbinya1996discrete} also proposed using WT for speech recognition.

Finally, considering the desire to demonstrate continuous GW analysis, further improvement of the optical microphone could lead to more advanced demonstration of techniques such as aiming or locking two or more optical microphones in an array.

\begin{figure*}
	\includegraphics[height=.99\textwidth, angle=-90]{figures/pipeline.pdf}
	\caption{
Flowchart for the signal through the optical microphone. The transfer function would have to account for each stage. 
The stage inside the red-dash box is analysed in Appendix~\ref{app:intensity_derivation}. Understanding the other stages is one aim of future work. 
}
	\label{fig:pipeline_highlighted}
\end{figure*}

### 2020-10-01 paper/ifo-wanderingTone.tex
Deeksha's comments - last few sections

Continuous wave searches, such as in Refs.~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017,SearchTwoSpecS6:2017,SunEtAlSNR:2018,JonesSun:2020}, are often (not always) performed on long datasets (months to years in duration).
The frequency of a continuous wave signals may wander significantly over this timescale, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or variable accretion flows onto a neutron star from a stellar companion (e.g. as in LMXBs)~\cite{BildstenTB:1998}.
In this context, wandering ``significantly'' means ``across multiple frequency bins'', where the typical width of a frequency bin is the reciprocal of the total observing time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
The audio analogue to these continuous wave signals is a tone that wanders in frequency; a note that changes pitch.
Here, we adapt the data analysis technique used to search for slowly wandering continuous waves signals from spinning neutron stars in aLIGO and aVirgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.

One method to search for a wandering signal is to split the time series into several shorter segments, analyse these individually, then recombine them (under some continuity assumption) to coherently analyse the whole.
The detection statistics in every segment are combined to form a grid in time and frequency (spectrogram).
The value at each grid point gives the likelihood that a signal is present at that particular time and frequency.
In gravitational-wave data analysis, a detection statistic is used to calculate the likelihood of a signal given the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun~\cite{JKS:1998}.
The detection statistic used depends on the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.

In this work the detection statistic is the amplitude (modulus) of the Fourier transform of the data at each time segment, which maximizes the likelihood of detecting a sinusoid in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}.
The resulting spectrogram gives the Fourier amplitude for each frequency and time bin.
In continuous wave searches a hidden Markov model is used to describe the system.
The frequency (the state) of the signal is unknown (hidden) and the signal is Markovian in that the state at any time depends only on the state at the previous time step.
The hidden Markov model method relates the observable quantity (the detection statistic) to the signal frequency (the hidden state).
The task is to find the most likely wandering frequency signal given the observable of the spectrogram.
The Viterbi algorithm~\cite{Viterbi:1967} is an efficient method to find the most likely path of a signal given an observation sequence.

This method is used in continuous wave searches for a range of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

The spectrogram grid is indexed by time $t$ and frequency $f$ where there are $N_t$ and $N_f$ time and frequency bins respectively, as shown in Fig.~\ref{fig:viterbi}.
We label the elements of the grid as $t_i$ and $f_j$ with $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. 
The normalised Fourier amplitude $F(t_i,f_j)$ is the detection statistic indicating the likelihood that a signal is present in each grid element.

The objective is to find the most likely path through the grid given the data and any probabilistic constraints on how the frequency wanders from $t_i$ to $t_{i+1}$. 
In continuous wave searches, a physical model is used to decide how much the frequency of the signal is allowed to wander at each time step, which depends on the type of target. 
In this work, at each time step we allow the frequency to either a) stay in the same bin; b) move up a single frequency bin; or c) move down a single frequency bin. 
In a hidden Markov model, the probability of transitioning from $f_j$ at $t_i$ to any of $\{f_{j+1},f_j,f_{j-1}\}$ at $t_{i+1}$ is called the transition probability $A(f_k,f_m)$. 
In this work the three options have equal probability $A(f_k,f_m)=1/3$ for $k=m+1,m,m-1$ and $A(f_k,f_m)=0$ otherwise.

A path through the grid has a likelihood dependent on the detection statistic (the observation) and the transition probability.
If the starting probabilities are all equal, i.e. ${\rm Pr}[f(t_0)] = 1/N_f$, the the probability of the frequency equalling $f(t_{i+1})$ at $t = t_{i+1}$ is defined recursively by
\begin{eqnarray}
{\rm Pr}[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &\times A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     &\times {\rm Pr}[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes ${\rm Pr}[f(t_{N_t}) = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

The Viterbi algorithm is an efficient method to find the Viterbi path. 
Here we describe the algorithm as pseudo-code referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the elements of the spectrogram grid (lighter colours correspond to higher likelihood). 
The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at $t_1$, each $f_j$ element has three possible paths it could originate from at $t_0$, indicated by the lines in Fig.~\ref{fig:viterbi}. 
At each $f_j$ element, the highest $F[t_0,f_j] A[f(t_0),f(t_1)]$ from the previous time step is selected, highlighted by the black lines in Fig.~\ref{fig:viterbi}. 
For example for the element labelled (a), the highest value is directly behind it, therefore the transition from $\{t_0,f_0\}$ to $\{t_1,f_0\}$ is selected. 
Each element has now selected the best path from the previous time step. 
In Fig.~\ref{fig:viterbi}, black lines show the best paths and grey lines show all other paths.

\item Moving to $t_2$, the path with the highest value of the recursive relation in Eqn.~\ref{eqn:recursiveViterbi} is selected for each $f_j$ as shown by the solid lines from $t_1$ to $t_2$ in Fig.~\ref{fig:viterbi}. 
The origins are stored for each grid point.

\item Steps 2 is repeated until the end of the grid ($t=t_{N_t}$) is reached. 
Only the best paths are stored.

\item At $t=t_{N_t}$, each $f_j$ has a value ${\rm Pr}[f(t_{N_t})=f_j]$ (Eqn.~\ref{eqn:recursiveViterbi}). 
The algorithm selects the terminating frequency $f(t_{N_t})$ with the highest value, marked (b) in Fig.~\ref{fig:viterbi}.

\item The Viterbi path is then found by backtracking along the stored best paths at each time step (see also Appendix~\ref{app:viterbi}). 
% come back to this. 
%\jam{[Andrew mentioned providing a closed form here, which I'm not sure exists. The code works by storing three arrays: the weights, the score for the best path to a given vertex, and the index of the previous vertex along that path. Backtracking comprises of looping through that third array from the vertex in the last row with the highest score from the second array. I don't think we need to explain it any further than that.]}
In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c). 
The Viterbi path is highlighted in orange in Fig.~\ref{fig:viterbi}.
\end{enumerate}
The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

To test the Viterbi algorithm, a sinusoidal signal with wandering frequency was played through the speaker and the output recorded via webcam as in Section~\ref{sec:single_tone}. 
The injected frequency itself followed a sinusoid with decaying amplitude. 
This choice was arbitrary, any injected path would have worked, as long as it changed slowly enough for the search parameters of the algorithm.

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. 
The overlaid pink-dot markers show the injected signal. 
The white-cross markers show the recovered Viterbi path maximised over all terminating frequency bins, which qualitatively provides a good match to the injection. 
Initially, the injected signal moves faster than the frequency wander limit, leading to the discrepancy observed for $t\lesssim 10\,{\rm s}$. 
There is also an anomaly at $150\,{\rm s}$, which illustrates what happens when the interferometer is disturbed, e.g. when someone walks nearby. 
The Viterbi algorithm is robust against such disturbances, recovering the optimal path promptly, as Fig.~\ref{fig:viterbi_overlay} shows.

Quantitatively, the Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. 
Another figure of merit is the root mean square (RMS) fractional error $E_{\rm rms}$ along the path defined as 
\han{}
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. 
Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay} we find $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

>>>>

Continuous wave searches, such as in Refs.~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017,SearchTwoSpecS6:2017,SunEtAlSNR:2018,JonesSun:2020}, are often (but not always) performed on long datasets (months to years in duration). The frequency of a continuous wave signals may wander significantly over this timescale, due to stochastic internal processes in the superfluid interior of isolated neutron stars~\cite{MelatosDouglassSimula:2015,Jones:2010} or variable accretion flows onto a neutron star from a stellar companion (e.g. as in LMXBs)~\cite{BildstenTB:1998}. In this context, wandering ``significantly'' means ``across multiple frequency bins'', where the typical width of a frequency bin is the reciprocal of the total observing time~\cite{JKS:1998,ScoX1O2Viterbi:2019}.
The audio analogue to these continuous wave signals is a tone that wanders in frequency; a note that changes pitch. Here, we adapt the data analysis technique used to search for slowly wandering continuous waves signals from spinning neutron stars in aLIGO and aVirgo data~\cite{SuvorovaEtAl:2016,SuvorovaEtAl:2017}.

One method to search for a wandering signal is to split the time series data into several shorter segments, analyse these individually, and then recombine them under some continuity assumption to coherently analyse the whole set. Most continuous wave searches use detection statistics to quantify the likelihood of a signal being present in the data at each frequency for each time series segment. The detection statistic used depends on the antenna beam pattern of the detector, which varies as the Earth rotates and orbits the Sun, and the type of target~\cite{JKS:1998,SuvorovaEtAl:2017}.
In this work, we take the discrete Fourier transform of each time series segment and use its amplitude at each frequency as a detection statistic. This ensures that we maximize the likelihood of detecting a sinusoidal signal in Gaussian noise, as described in Appendix~\ref{app:sinusoid_likelihood}. The likelihoods (Fourier amplitudes) from each segment are combined to form a grid in time and frequency (a spectrogram). We also normalise the detection statistics for convenience.

We use a hidden Markov model (HMM) to recover the wandering signal from noisy data. To do this, we define the frequency of the signal as a hidden (unknown) state which can undergo transitions at discrete times. We assume that these transitions are Markovian in that the hidden state (i.e.\ frequency) of the system at any time depends solely on its state at the previous time step. The detection statistic (observable) then relates the hidden states of the system to the observed data.
To perform the task of recovering the signal, we use the Viterbi algorithm~\cite{Viterbi:1967} to efficiently find the most probable path through the spectrogram given the sequence of observations.
This method is used in continuous wave searches for a variety of astrophysical targets~\cite{ScoX1O2Viterbi:2019,ScoX1ViterbiO1:2017,MillhouseStrangMelatos:2020,JonesSun:2020,MiddletonEtAlO2LMXBs:2020,PostMergerRemnantSearch:2019,SunEtAlSNR:2018,viterbi_application}.

@@

The spectrogram, as shown in Fig.~\ref{fig:viterbi}, has $N_t$ time and $N_f$ frequency bins. From here on, we will call these time “steps” instead of time bins, but they should not be confused with the timesteps of the original time series. We label each time and frequency step as $t_i$ and $f_j$ with $i=0,1,2,...N_t$ and $j=0,1,2,...,N_f$. At each state (grid point) the detection statistic (the normalised Fourier amplitude) is notated $F(t_i,f_j)$.

The objective is to find the most likely path through the grid given the observed data and any probabilistic constraints on how the frequency of the signal can wander between time steps $t_i$ to $t_{i+1}$. In continuous wave searches, a physical model of the target informs how far the frequency of the signal can wander over each time step. This information is encoded in the transition probability matrix $A(f_k,f_m)$, which describes the probability of system transitioning from state $f_i$ at $t_i$ to a state $f_{j}$ at $t_{i+1}$. Here, we allow the frequency of the signal to either (a) stay in the same bin; (b) move up by a single frequency bin; or (c) move down by a single frequency bin at each time step. We assign these three transitions equal probability, i.e.\ $A(f_k,f_m)=1/3$ for $k=m+1,m,m-1$ and $A(f_k,f_m)=0$ otherwise.

We define the probability of the system having frequency $f_j$ at the initial time $t_0$ to be equal to the (normalised) detection statistic at that state (i.e.\  ${\rm Pr}[f(t_0) = f_j] = F[t_0,f_j]$).
The probability of the system having frequency $f_j$ at any later time $t_{i+1}$ is then defined recursively by
\begin{eqnarray}
{\rm Pr}[f(t_{i+1})=f_j] =~& F[t_{i+1},f(t_{i+1})] \nonumber \\
                     &\times A[f(t_{i+1}),f(t_i)]  \nonumber \\
                     &\times {\rm Pr}[f(t_i)].
\label{eqn:recursiveViterbi}
\end{eqnarray}
The sequence $f(t_0),f(t_1),\dots,f(t_{N_t})$ that maximizes ${\rm Pr}[f(t_{N_t}) = f_j]$ is the optimal Viterbi path terminating in the frequency bin $f_j$. 
We can the maximize the latter quantity over $0 \leq j \leq N_f$ to find the optimal Viterbi path overall, i.e. terminating in any frequency bin.

The Viterbi algorithm provides a computationally efficient method for finding the optimal path, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}). Here we describe the algorithm in pseudocode while referring to the schematic in Fig.~\ref{fig:viterbi}, where the circles indicate the states in the spectrogram with the lighter colours corresponding to a higher likelihood. The implementation used in this work is available online (see Appendix~\ref{app:code}).
\begin{enumerate}
\item Starting at time $t_1$, each $f_j$ state can originate from three prior states at time $t_0$ (except for $f_0$ and $f_{N_f}$ which only have two). The paths between these states are indicated by the lines in Fig.~\ref{fig:viterbi}. At each $f_j$ state, we select the path with the highest $A[f(t_0),f(t_1)] {\rm Pr}[f(t_0)]$ as the most probable path. These choices are highlighted using the black lines in Fig.~\ref{fig:viterbi} while the grey lines show the rejected paths. For example, the most probable connection to the element labelled (a) is the one directly behind it (i.e., $f_4$). Therefore, this path is selected as the best path from $t_0$ to $t_1$. For backtracking at the end, the index of the most probable connection along with the current cost of the best path is stored for each node.

\item Moving to the third time step, $t_2$, again, we select the path which maximises the recursive quantity in Eqn.~\ref{eqn:recursiveViterbi} for each $f_j$. These are again shown by the solid black lines between the nodes at $t_1$ and $t_2$ in Fig.~\ref{fig:viterbi}.

\item Step 2 is repeated until the end of the grid ($t=t_{N_t}$) is reached with only the best paths being stored at each iteration.

\item Each state $f_j$ at $t=t_{N_t}$ has some value ${\rm Pr}[f(t_{N_t})=f_j]$ (Eqn.~\ref{eqn:recursiveViterbi}), which represents the probability of the most likely path that ends in state $f_j$ at $t_{N_t}$. The Viterbi algorithm then selects the terminating frequency bin $f(t_{N_t})$ with the highest probability, marked by (b) in Fig.~\ref{fig:viterbi}.

\item The Viterbi path (the overall best path) is then found by backtracking along the stored best connections at each time step (see also Appendix~\ref{app:viterbi}). In Fig.~\ref{fig:viterbi}, we see that the path ending at (b) started at (c) and is highlighted in orange in Fig.~\ref{fig:viterbi}.
\end{enumerate}

In the following section, we use the Viterbi algorithm to recover a slowly wandering signal.

% The Viterbi algorithm is an example of a dynamical programming algorithm, where a computation can be broken down into a series of sub-computations.
% The Viterbi path for a time series sequences spanning $t_0$ to $t_{N_t}$ contains the Viterbi path for sub-sequences of that time series. 
% The algorithm is computationally efficient, at every time step all but $N_f$ possible paths are eliminated (see also Ref.~\cite{ScoX1ViterbiO1:2017}).

@@

% throughout this section we are using present tense, should this be changed to past tense?

A sinusoidal signal with a slowly decaying amplitude is injected into the interferometer using the speaker in the setup shown in Fig~\ref{fig:ifo_schematic_webcam} to mimic a slowly wandering signal. The particular choice of signal is arbitrary as any injected path would have worked, as long as it changed slowly enough for the search parameters of the algorithm. The output of the interferometer is recorded via a webcam as in Section~\ref{sec:single_tone}. We then test the Viterbi algorithm’s ability to recover the wandering signal.

The results are shown in Fig.~\ref{fig:viterbi_overlay}, where the heatmap shows the spectrogram of the observed signal. The overlaid pink dots show the injected signal while the white crosses show the recovered Viterbi path (the most likely path through the data according to the Viterbi algorithm). The recovered Viterbi path is within one frequency bin ($\approx 0.3\,{\rm Hz}$) of the injected signal for $94\%$ of the time. We also compute the root mean square (RMS) fractional error $E_{\rm rms}$ along the path which is defined as 
\begin{equation}
E_{\rm rms} = \left[ \frac{1}{N_t+1} \sum_{i=0}^{N_t} \frac{(I_i - R_i)^2}{{I_i}^2} \right] ^{1/2}
\end{equation}
where $I_i$ and $R_i$ are the injected and recovered (Viterbi) frequency paths. Computing this for the result shown in Fig.~\ref{fig:viterbi_overlay}, we find that $E_{\rm rms} = 0.082$, which indicates fair but not total recovery of the injected signal.

This may be explained by two anomalies in the recovered path. Initially, the injected signal wanders by more than one frequency bin at each time step (i.e.\ faster than the algorithm is allowed to), thus leading to a discrepancy between the injected and recovered paths for $t\lesssim 10\,{\rm s}$. One may be tempted to increase the allowed frequency wander in the algorithm, however this leads to an overall decrease in the above statistics, as the algorithm is prone to jump briefly to nearby spots of noise. There is also an anomaly at $150\,{\rm s}$, which is likely due to a local disturbance, e.g. someone walked past the interferometer. The Viterbi algorithm is somewhat robust against such disturbances, qualitatively recovering the injected path on the other side, as shown in Fig.~\ref{fig:viterbi_overlay}.
